# SummarizedExperiments {#sec-se}

```{r, echo = FALSE, message = FALSE}
library("tidyverse")
```

**Objectives**

- Present the `SummarizedExperiment` class, extensively used in omics
  analyses.
- Show how to build `SummarizedExperiment` objects.

Data in bioinformatics is often more complex than the basic data types
we have seen so far. To deal with this, developers define specialised
data containers (termed classes) that match the properties of the data
they need to handle.

This aspect is central to the **Bioconductor**[^Bioconductor] project
which uses the same **core data infrastructure** across packages. This
certainly contributed to Bioconductor's success. Bioconductor package
developers are advised to make use of [existing
infrastructure](https://uclouvain-cbio.github.io/WSBIM1322/sec-obj.html#examples-from-the-bioconductor-project)
to provide coherence, interoperability and stability to the project as
a whole.

[^Bioconductor]:The [Bioconductor](http://www.bioconductor.org) was
initiated by Robert Gentleman, one of the two creators of the R
language. Bioconductor provides tools for the analysis and
comprehension of omics data.  Bioconductor uses the R statistical
programming language, and is open source and open development.

To illustrate such an omics data container, we'll present the
`SummarizedExperiment` class.


## SummarizedExperiment

The figure below represents the anatomy of the SummarizedExperiment class.

```{r SE,  echo=FALSE, out.width = '80%'}
knitr::include_graphics("https://uclouvain-cbio.github.io/WSBIM1322/figs/SE.svg")
```

Objects of the class SummarizedExperiment contain :

- **One (or more) assay(s)** containing the quantitative omics data
  (expression data), stored as a matrix-like object. Features (genes,
  transcripts, proteins, ...) are defined along the rows, and samples
  along the columns.

- A **sample metadata** slot containing sample co-variates, stored as a
  data frame. Rows from this table represent samples (rows match exactly the
  columns of the expression data).

- A **feature metadata** slot containing feature co-variates, stored as
  a data frame. The rows of this data frame match exactly the rows of the
  expression data.

The coordinated nature of the SummarizedExperiment guarantees that
during data manipulation, the dimensions of the different slots will
always match (i.e the columns in the expression data and then rows in
the sample metadata, as well as the rows in the expression data and
feature metadata) during data manipulation. For example, if we had to
exclude one sample from the assay, it would be automatically removed
from the sample metadata in the same operation.

The metadata slots can grow additional co-variates
(columns) without affecting the other structures.


### Creating a SummarizedExperiment

In order to create a `SummarizedExperiment`, we will create the
individual components, i.e the count matrix, the sample and gene
metadata from csv files. These are typically how RNA-Seq data are
provided (after raw data have been processed).

```{r eval=FALSE, message=FALSE, include=FALSE}
# Simplify dataset
library(org.Hs.eg.db)
load(file = "wsbim2122_data/deseq2/counts_both.rda")
counts <- as_tibble(counts, rownames = "ENSEMBL") |>
  mutate(gene = unname(mapIds(org.Hs.eg.db, ENSEMBL, "SYMBOL", "ENSEMBL"))) |>
  filter(!is.na(gene)) |>
  filter(!duplicated(gene)) |>
  dplyr::select(gene, starts_with("sample"))

load(file = "wsbim2122_data/deseq2/coldata_both.rda")
coldata <- as_tibble(coldata, rownames = "sample") |>
    dplyr::select(-Cell)
coldata$Condition[coldata$Condition == "mock"] <- "ctl"
coldata$sample <- c(paste0("Epi_ctl", 1:3), paste0("Epi_KD", 1:3),
                    paste0("Fib_ctl", 1:3), paste0("Fib_KD", 1:3))

names(counts) <- c("gene", coldata$sample)

rowdata <- tibble(gene = counts$gene,
                  ENSEMBL = mapIds(org.Hs.eg.db, gene, "ENSEMBL", "SYMBOL"),
                  chr = mapIds(org.Hs.eg.db, ENSEMBL, "MAP", "ENSEMBL")) |>
  mutate(chr = sub("p.*", "", x = chr)) |>
  mutate(chr = sub("q.*", "", x = chr))


dir.create("data", showWarnings = F)
write_csv(counts, file = "data/count_matrix.csv")
write_csv(coldata, file = "data/sample_metadata.csv")
write_csv(rowdata, file = "data/gene_metadata.csv")
```



- **An expression matrix**: we load the count matrix, specifying that
  the first columns contains row/gene names, and convert the
  `data.frame` to a `matrix`. You can download it
  [here](https://raw.githubusercontent.com/UCLouvain-BIOINFO/bioinfo-training-03-rnaseq/refs/heads/main/data/count_matrix.csv).


```{r countMatrix}
count_matrix <- read.csv("data/count_matrix.csv",
                         row.names = 1) |>
    as.matrix()
count_matrix[1:5, ]
dim(count_matrix)
```


- **A table describing the samples**, available
  [here](https://raw.githubusercontent.com/UCLouvain-BIOINFO/bioinfo-training-03-rnaseq/refs/heads/main/data/gene_metadata.csv).
  

```{r sampleMetadata, message = FALSE}
sample_metadata <- read_csv("data/sample_metadata.csv")
sample_metadata
dim(sample_metadata)
```

- **A table describing the genes**, available
  [here](https://raw.githubusercontent.com/UCLouvain-BIOINFO/bioinfo-training-03-rnaseq/refs/heads/main/data/sample_metadata.csv).

```{r geneMetadata, message = FALSE}
gene_metadata <- read_csv("data/gene_metadata.csv")
head(gene_metadata)
dim(gene_metadata)
```

We will create a `SummarizedExperiment` from these tables:

- The count matrix that will be used as the **`assay`**

- The table describing the samples will be used as the **sample
  metadata** slot

- The table describing the genes will be used as the **features
  metadata** slot

To do this we can put the different parts together using the
`SummarizedExperiment` constructor:

```{r, message = FALSE}
## BiocManager::install("SummarizedExperiment")
library("SummarizedExperiment")
```

First, we make sure that the samples are in the same order in the
count matrix and the sample annotation, and the same for the genes in
the count matrix and the gene annotation.

```{r}
stopifnot(rownames(count_matrix) == gene_metadata$gene)
stopifnot(colnames(count_matrix) == sample_metadata$sample)
```


```{r}
se <- SummarizedExperiment(assays = list(counts = count_matrix),
                           colData = sample_metadata,
                           rowData = gene_metadata)
se
```

Using this data structure, we can access the expression matrix with
the `assay` function:

```{r}
head(assay(se))
dim(assay(se))
```

We can access the sample metadata using the `colData` function:

```{r}
colData(se)
dim(colData(se))
```

We can also access the feature metadata using the `rowData` function:

```{r}
head(rowData(se))
dim(rowData(se))
```

### Subsetting a SummarizedExperiment

SummarizedExperiment can be subset just like with data frames, with
numerics or with characters of logicals.

Below, we create a new instance of class SummarizedExperiment that
contains only the 5 first features for the 3 first samples.

```{r}
se1 <- se[1:5, 1:3]
se1
```

```{r}
colData(se1)
rowData(se1)
```


We can also use the `colData()` function to subset on something from
the sample metadata or the `rowData()` to subset on something from the
feature metadata.

#### Question {-}

Create a new object keeping only fibroblast samples.

<details>
<summary>Solution</summary>

```{r}
se1 <- se[, colData(se)$Type == "Fibroblast"]

# Equivalent to
se1 <- se[, se$Type == "Fibroblast"]

dim(se1)
colData(se1)
```

</details>

#### Question {-}

Extract the gene expression levels of genes located on sex chromosomes 
in Fibroblast control cells

<details>
<summary>Solution</summary>

```{r, purl=FALSE}
se2 <- se[rowData(se)$chr %in% c("X", "Y"), 
          se$Type == "Fibroblast" & se$Condition == "ctl"]
head(assay(se2))
rowData(se2)
```

</details>

### Adding variables to the sample metadata

We can also add information to the sample metadata. Suppose that you want to
add the center where the samples were collected ...

```{r}
colData(se)$university <- rep("UCLouvain", nrow(colData(se)))
colData(se)
```

This illustrates that the metadata slot can grow indefinitely without
affecting the other structures!


#### Question {-}

Create a "group" column to to the metadata (i.e "Epithelial_ctl", "Epithelial_KD", ...)

<details>
<summary>Solution</summary>

```{r}
se$group <- paste(se$Type, se$Condition, sep = "_")
```

</details>

### Adding gene annotations to the feature metadata

The rowData slot of our se provides the ENSEMBL gene references. It might be 
interesting to add other types of informative annotations describing genes.

#### Gene annotation database

`AnnotationDbi` is an R package that provides an interface for connecting
and querying various annotation databases.

There are a many organism-specific `orgDb` packages, such as `org.Hs.eg.db`
for human and `org.Mm.eg.db` for mouse.

```{r message = FALSE}
#BiocManager::install(c("AnnotationDbi", "org.Hs.eg.db"))
library("AnnotationDbi")
library("org.Hs.eg.db")
```

The `keytypes()` function can help us to see what information we can extract
from the database (ENSEMBL gene ID, gene symbol, gene name, ...)


```{r}
keytypes(org.Hs.eg.db)
```

We can easily extract information from this database using the `mapIds()` function.

For example, we can retrieve the ENTREZID references, or the chromosomal 
location of the 5 first genes as follow:


```{r}
# Retrieving the ENTREZID ids
mapIds(x = org.Hs.eg.db,
       keys = rownames(se)[1:5],
       column = "ENTREZID",
       keytype = "SYMBOL")

# Retrieving the chromosomal location 
mapIds(x = org.Hs.eg.db,
       keys = rownames(se)[1:5],
       column = "MAP",
       keytype = "SYMBOL")
```


#### Questions {-}

1. Add gene biotypes (stored as `GENETYPE` in `org.Hs.eg.db` object) to the 
feature metadata slot of our SummarizedExperiment se.

2. How many genes from each biotype do we have?

<details>
<summary>Solutions</summary>

1. Add gene biotypes (stored as `GENETYPE` in `org.Hs.eg.db` object) to the 
feature metadata slot of our SummarizedExperiment se.

```{r, purl=FALSE}
rowData(se)$gene_biotype <-  mapIds(x = org.Hs.eg.db, keys = rownames(se), 
                                    column = "GENETYPE", keytype = "SYMBOL")
head(rowData(se))
```


2. How many protein-coding genes do we have?

```{r, purl=FALSE}
se[rowData(se)$gene_biotype == "protein-coding",]
```

</details>




### Save a SummarizedExperiment

This was a bit of code and time to create our `SummarizedExperiment` object.
We will need to keep using it throughout the workshop, so it can be useful to
save it as an actual single file on our computer to read it back in to R's memory
if we have to shut down RStudio. To save an R-specific file we can use the `saveRDS()`
function and later read it back into R using the `readRDS()` function.

```{r save-se, purl=TRUE}
saveRDS(se, "data/se.rds")
rm(se) # remove the object!
se <- readRDS("data/se.rds")
```




## RangedSummarizedExperiment

In addition to the `SummarizedExperiment` slots we have seen above, it
is also possible to store information about the features' genomic
ranges. Below, we load the `airway` `RangedSummarizedExperiment` data
to illustrate this:


```{r load_airway}
library(airway)
data(airway)
airway
```

```{r rowRanges}
rowRanges(airway)
```


This `GRangesList` contains a list of 64102 range elements, one for
each gene/feature. Each element is a `GRanges` object that contains
the positions of the exons or the genes:


```{r rowRanges1}
rowRanges(airway)[[1]]
```

It is possible to use these row genomic ranges to subset features in
regions of interest, defined as `GRanges` object. Below, for example,
we subset features between bases 1M and 2M on chromosome 7:


```{r subsetByOverlaps}
roi <- GRanges(seqnames = "7", ranges = 1e6:2e6)
subsetByOverlaps(airway, roi)
```
