# Enrichment analyses {#sec-gsea}


**Objectives**


- What is the aim of performing gene set enrichment analysis?
- What are the commonly-used gene set databases?
- Learn how to perform ORA and GSEA analyses


## Objectives of an enrichment analysis

Once we have obtained a list of differentially expressed (DE) genes,
the next question naturally to ask is what biological functions these
DE genes may affect.

Gene set enrichment analyses (GSEA) evaluate the associations of a
list of DE genes to a collection of pre-defined gene sets, where each
gene set has a specific biological meaning. A gene set significantly
enrichiched among DE genes could suggest that the corresponding
biological process or pathway is significantly affected.

There are a huge amount of methods available for enrichment
analyses. In this chapter, we will focus on two widely used methods:
the over-representation analysis (ORA) and the Gene Set Enrichment
Analysis (GSEA).

```{r load-pkg-enrich, include=FALSE}
suppressPackageStartupMessages({
  library(DESeq2)
  library(tidyverse)
  library(org.Hs.eg.db)
  library(msigdbr)
  library(clusterProfiler)})
```


## Gene sets

The definition of a gene set is very flexible and the construction of gene
sets is straightforward. In most cases, gene sets are from public databases
where huge efforts from scientific curators have already been made to
carefully categorise genes into gene sets with clear biological meanings.
For example, in a “cell cycle” gene set, all the genes are involved in the cell
cycle process. Thus, if DE genes are significantly enriched in the “cell cycle”
gene set, which means there are significantly more cell cycle genes differentially
expressed than expected, we can conclude that the normal function of cell cycle
process may be affected.

Nevertheless, gene sets can also be self-defined from individual studies, such
as a set of genes in a network module from a co-expression network analysis,
or a set of genes that are up-regulated in a certain disease.

The [MSigDB database](https://www.gsea-msigdb.org/gsea/msigdb) contains a collection
of annotated gene sets such as those presented below.

#### Gene Ontology (GO) {-}

Gene Ontology defines concepts/classes used to describe gene function,
and relationships between these concepts. GO terms are organized in a
directed acyclic graph, where edges between terms represent
parent-child relationship.  It classifies functions along three
aspects.

- **Molecular Function (MF)**: molecular activities of gene products
- **Cellular Component (CC)**: where gene products are active
- **Biological Process (BP)** pathways and larger processes made up of
  the activities of multiple gene products.

#### Hallmark gene sets {-}

Hallmark gene sets are curated gene sets that represent well-defined
biological states or processes (e.g. Apoptosis, KRAS signaling, DNA
repair).

#### Kyoto Encyclopedia of Genes and Genomes {-}

Kyoto Encyclopedia of Genes and Genomes (KEGG) is a collection of
manually drawn pathway maps representing molecular interaction and
reaction networks. These pathways cover a wide range of biochemical
processes.

#### Other gene sets {-}

Other gene sets include but are not limited to Positional gene sets,
wikiPathways, Oncogenic signatures, Immunologic signatures ...


## Access the gene sets from MSigDB

```{r getMsigdb, eval=FALSE, message=FALSE, include=FALSE}
## listCollections(MSigDBGeneSets)
## listSubCollections(MSigDBGeneSets)

MSigDBGeneSets <- getMsigdb(org = "mm", id = "SYM", version = "7.4")

GOBP <- subsetCollection(MSigDBGeneSets,
                         collection = "c5",
                         subcollection = "GO:BP")

## problem with function: the subcollection doesn't seem to work, all
## GO terms were selected + No way to retrieve KEGG pathways

gene_sets <- stack(geneIds(GOBP)) |> #stack() transforms a list of elements into a dataframe
  as_tibble() |>
  dplyr::rename(gs_name = ind, gene = values) |>
  dplyr::select(gs_name, gene)

gene_sets
#=> Use the msigdbr instead
```

The
[msigdbr](https://cran.r-project.org/web/packages/msigdbr/index.html)
CRAN package provides the MSigDB gene sets in a standard R data frame
with key-value pairs.

One can use `msigdbr_collections()` to see the available collections:

```{r msigdbr-collections}
msigdbr_collections() |>
    as_tibble()
```

The `msigdbr()` function retrieves a data frame of gene sets and their
member genes. Below we set the species of interest, the GO collection
C5, and more specifically the Biological Process sub-collection. We
also select the gene set name and gene symbols among the 20 variables.

```{r GOBP}
GO_BP <- msigdbr(species = "Homo sapiens",
                 collection = "C5",
                 subcollection = "BP") |>
    dplyr::select(gs_name, gene_symbol)

GO_BP
```


#### Question {-}

How many genes belong to the `GOBP_REGULATION_OF_LYMPHOCYTE_MIGRATION`
geneset?

<details>
<summary>Solution</summary>


```{r GOBP_REGULATION_OF_LYMPHOCYTE_MIGRATION}
GO_BP |>
  filter(gs_name == "GOBP_REGULATION_OF_LYMPHOCYTE_MIGRATION")
```

</details>

#### Question {-}

Retrieve human `Hallmarks` gene sets. How many different gene sets are
available?

<details>
<summary>Solution</summary>


```{r hallmarks}
hallmarks <- msigdbr(species = "Homo sapiens",
                     collection = "H") |>
    dplyr::select(gs_name, gene_symbol)

hallmarks

table(hallmarks[[1]])
```

</details>


## Input data

We will use the results of the differential expression analysis
comparing KD to mock epithelial cells. The following code summarizes
the full `DESeq2` analysis that you have applied in the previous
chapter.  We will focus on the genes that have an adjusted p-value
(i.e. those that have been tested).

```{r enrich-quick-deseq2, message = FALSE, warning = FALSE}
se <- readRDS("data/se.rds")
dds <- DESeqDataSet(se, design = ~ group)
dds <- DESeq(dds)
res <- results(dds, name = "group_Epithelial_KD_vs_Epithelial_ctl")
res_tbl <- as_tibble(res, rownames = "gene")
res_tbl <- res_tbl |> filter(!is.na(padj))
```

## Over-representation analysis

### Principle

The idea behind the ORA is to test whether your list of differentially
expressed genes (DE) is enriched in certain biological categories
(such as gene sets from Gene Ontology terms, KEGG pathways, or
Reactome pathways), i.e, if it contains more genes from these
categories than would be expected by chance.

#### Step 1: selection of DE genes {-}

To perform an over representation analysis, we first need to define
the genes we will consider as differentially expressed (DE).

```{r ora-step-1, echo=FALSE, fig.height=6, fig.width=15, message=FALSE, warning=FALSE}
padj_thr <- 0.05
log2FC_thr <- 0
genes_DE1 <- res_tbl$gene[res_tbl$padj <= padj_thr &
                  abs(res_tbl$log2FoldChange) >= log2FC_thr]

fig1 <- res_tbl |>
  ggplot(aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = gene %in% genes_DE1), size = 1) +
  scale_color_manual(values = c("gray", "red")) +
  theme_bw() +
  theme(legend.position = "none") +
  geom_hline(yintercept = -log10(padj_thr)) +
  geom_vline(xintercept = log2FC_thr) +
  ggtitle(paste0(length(genes_DE1), " DE selected (padj < ", padj_thr, ")"))


padj_thr <- 1e-10
log2FC_thr <- 0
genes_DE2 <- res_tbl$gene[res_tbl$padj <= padj_thr &
                  abs(res_tbl$log2FoldChange) >= log2FC_thr]

fig2 <- res_tbl |>
  ggplot(aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = gene %in% genes_DE2), size = 1) +
  scale_color_manual(values = c("gray", "red")) +
  theme_bw() +
  theme(legend.position = "none") +
  geom_hline(yintercept = -log10(padj_thr)) +
  geom_vline(xintercept = log2FC_thr) +
  ggtitle(paste0(length(genes_DE2), " DE selected (padj < ", padj_thr, ")"))

padj_thr <- 1e-10
log2FC_thr <- 1
genes_DE3 <- res_tbl$gene[res_tbl$padj <= padj_thr &
                  abs(res_tbl$log2FoldChange) >= log2FC_thr]

fig3 <- res_tbl |>
  ggplot(aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = gene %in% genes_DE3), size = 1) +
  scale_color_manual(values = c("gray", "red")) +
  theme_bw() +
  theme(legend.position = "none") +
  geom_hline(yintercept = -log10(padj_thr)) +
  geom_vline(xintercept = log2FC_thr) +
  geom_vline(xintercept = -log2FC_thr) +
  ggtitle(paste0(length(genes_DE3), " DE selected (padj < ", padj_thr,
                 ' and |logFC| > ', log2FC_thr, ")"))

fig1 + fig2 + fig3
```

We could consider all genes that have a p-adjusted value < 0.05
(left), in which case we would select `r length(genes_DE1)`
genes. This number is probably too large, so we should probably be
more restrictive and select a smaller set of genes, using a more
stringent filtering on the p-adjusted value (center).  We could also
consider filtering using the logFC, which could also make biological
sense (right).

Whatever thresholds are used, the idea here is to start from a
selection of genes that we will consider as our DE genes.  **The
selection is a choice the user has to make.**

Here, we will apply the selection criteria illustrated in the second
volcano. Hence, among all the genes in our experiment (called the
**universe**), we will consider the `r length(genes_DE2)` genes with a
p-adjusted value < `r padj_thr` as our **genes of interest** or **DE
genes**.

#### Step 2: Counting {-}

For each biological category, we can count how many genes from our DE
genes are in that category, and how many are not.

```{r my-geneset}
my_geneset <- "GOBP_ENDODERM_DEVELOPMENT"
```

Let's say we want to test the enrichment of the `r my_geneset` gene
set. The volcano plot below highlights the genes from the set in blue.

```{r ora-GOBP-ENDODERM-DEVELOPMENT, echo=FALSE, out.width='80%'}
GO_BP <- msigdbr(species = "Homo sapiens",
                 collection = "C5",
                 subcollection = "BP") |>
    dplyr::select(gs_name, gene_symbol)

genes_from_geneset <- GO_BP |>
    filter(gs_name == my_geneset) |>
    pull(gene_symbol)

padj_thr <- 1e-10
log2FC_thr <- 0

res_tbl |>
    ggplot(aes(x = log2FoldChange, y = -log10(padj))) +
    geom_point(color = "gray70", size = 1) +
    geom_point(data = res_tbl |> filter(gene %in% genes_from_geneset),
               color = "blue", size = 1) +
    theme(legend.position = "none") +
    geom_hline(yintercept = -log10(padj_thr)) +
    geom_vline(xintercept = log2FC_thr) +
    geom_vline(xintercept = -log2FC_thr) +
    ggtitle(my_geneset) +
    theme_bw() +
    theme(title = element_text(color = "blue"),
          axis.title = element_text(color = "black"),
          aspect.ratio = 1)
```


We will count the number of DE genes from that are in
`r my_geneset` gene set, and how many are not.

```{r enrich-ora-table, echo=FALSE, fig.align='center'}
# DE and GO
n <- length(intersect(res_tbl$gene[res_tbl$gene %in% genes_DE2],
                      genes_from_geneset))

# not DE and GO
m <- length(intersect(res_tbl$gene[!res_tbl$gene %in% genes_DE2],
                      genes_from_geneset))

# DE and not GO
p <- length(setdiff(res_tbl$gene[res_tbl$gene %in% genes_DE2],
                      genes_from_geneset))

# not DE and not GO
q <- length(setdiff(res_tbl$gene[!res_tbl$gene %in% genes_DE2],
                      genes_from_geneset))

count_mat <- matrix(c(n, m, p, q), nrow = 2)
rownames(count_mat) <- c("DE", "not_DE")
colnames(count_mat) <- c("GO", "not_GO")

knitr::kable(count_mat)
```


#### Step 3: Statistical test {-}

We can now apply a Fisher’s exact (or hypergeometric test) that will
test whether there is a significant enrichment of DE genes in the GO
term.

```{r enrich-fisher-test}
fisher.test(count_mat, alternative = "greater")
```
### Running an ORA with the `clusterProfiler` package

In practice, the analysis presented above can be executed using any of
the very many packages that are available.

Here, we will use the [clusterProfiler
package](https://yulab-smu.top/clusterProfiler-book/) to test if any
GO terms from the Biological Process is enriched among our DE genes.

Let's start be defining our DE genes and our universe.

```{r ora-de-genes}
padj_thr <- 1e-10
log2FC_thr <- 0

genes_DE <- res_tbl |>
    filter(padj <= padj_thr,
           abs(log2FoldChange) >= log2FC_thr) |>
    pull(gene)

geneList <- res_tbl$gene
```

We won't set any (adjusted) p-value threshold and filter the results
by hand to be able to adapt the it without re-running the analysis.

```{r run-ora}
library(clusterProfiler)
ORA_res <- enricher(gene          = genes_DE,
                    universe      = geneList,
                    pAdjustMethod = "BH",
                    qvalueCutoff  = 1,
                    pvalueCutoff  = 1,
                    TERM2GENE     = GO_BP) |>
    as_tibble()

ORA_res_tbl_GO <- ORA_res |>
    filter(p.adjust < 0.05)

ORA_res_tbl_GO
```

In the output data frame, there are the following columns:

- `ID`: ID of the gene set. In this example analysis, it is the GO ID.
- `Description`: Readable description. Here it is the name of the GO term.
- `GeneRatio`: Number of DE genes in the gene set / total number of DE genes.
- `BgRatio`: Size of the gene set / total number of genes.
- `pvalue`: _p_-value calculated from the hypergeometric distribution.
- `p.adjust`: Adjusted _p_-value by the BH method.
- `qvalue`: _q_-value which is another way for controlling false
  positives in multiple testings.
- `geneID`: A list of DE genes in the gene set.
- `Count`: Number of DE genes in the gene set.

You may have noticed the total number of DE genes changes. We defined
`r length(genes_DE)` DE genes, but only
`r length(genes_DE[genes_DE %in% GO_BP$gene_symbol])` DE genes are
included in the enrichment result table (in the GeneRatio column).
Similarly, our universe had `r length(geneList)` genes, but the total number of
tested genes was only `r nrow(res_tbl |> filter(gene %in% GO_BP$gene_symbol))`
in the result table. The main reason is by default DE genes not annotated to any
GO gene set are filtered out.

### Visualisation

The clusterProfiler documentation provides a chapter on the
[visualization of functional enrichment
results](https://yulab-smu.top/clusterProfiler-book/chapter12.html).

Another useful visualisation, that links the enrichment results back
to the whole set of results is to highlight the genes in a particular
set of interest on the volcano plot.

Let's inspect genes from the
`GOBP_RESPONSE_TO_TRANSFORMING_GROWTH_FACTOR_BETA`on the volcano


```{r volcano-GOBP_RESPONSE_TO_TRANSFORMING_GROWTH_FACTOR_BETA}
my_geneset <- "GOBP_RESPONSE_TO_TRANSFORMING_GROWTH_FACTOR_BETA"

genes_from_geneset <- GO_BP |>
  filter(gs_name == my_geneset) |>
  pull(gene_symbol)

padj_thr <- 1e-10
log2FC_thr <- 0

res_tbl |>
    ggplot(aes(x = log2FoldChange, y = -log10(padj))) +
    geom_point(color = "gray80", size = 1) +
    geom_point(data = res_tbl |>
                   filter(gene %in% genes_from_geneset),
               color = "purple", size = 1) +
    theme(legend.position = "none") +
    geom_hline(yintercept = -log10(padj_thr)) +
    geom_vline(xintercept = log2FC_thr) +
    geom_vline(xintercept = -log2FC_thr) +
    ggtitle(paste0(length(genes_DE2), " DE selected (padj < ", padj_thr, ")")) +
    ggtitle(my_geneset) +
    theme_bw() +
    theme(title = element_text(color = "purple"),
          axis.title = element_text(color = "black"),
          aspect.ratio = 1)
```


### Conclusion about ORA method

This approach is straightfoward and very fast. Its major drawback
however is that we need to define a cutoff to differentiate DE from
non-DE genes.  Setting this threshold might have an effect on the
results.


#### Question {-}

Test different p-value and/or log fold-change thresholds to define the
group of 'DE genes'.

<details>
<summary>Solution</summary>


```{r ora-2, warning = FALSE}
padj_thr <- 1e-10
log2FC_thr <- 1

genes_DE <- res_tbl |>
    filter(padj <= padj_thr,
           abs(log2FoldChange) >= log2FC_thr) |>
    pull(gene)

geneList <- res_tbl$gene

res_tbl |>
    ggplot(aes(x = log2FoldChange, y = -log10(padj))) +
    geom_point(aes(color = gene %in% genes_DE), size = 1) +
    scale_color_manual(values = c("gray", "red")) +
    theme_bw() +
    theme(legend.position = "none") +
    geom_hline(yintercept = -log10(padj_thr)) +
    geom_vline(xintercept = -log2FC_thr) +
    geom_vline(xintercept = log2FC_thr) +
    ggtitle(paste0(length(genes_DE), " DE selected (padj < ", padj_thr,
                   ", abs(lfc) > ", log2FC_thr, ")"))

ORA_res_new_thr <- enricher(gene          = genes_DE,
                            universe      = geneList,
                            pAdjustMethod = "BH",
                            qvalueCutoff  = 1,
                            pvalueCutoff  = 1,
                            TERM2GENE     = GO_BP) |>
    as_tibble()

ORA_res_new_thr |>
    filter(p.adjust < 0.05)

ORA_res_GO_new_thr <- ORA_res_new_thr |>
    filter(p.adjust < 0.1)
```

```{r venn-ora-ora, warning = FALSE}
if (!require(Vennerable))
    BiocManager::install("js229/Vennerable")

geneset_list <- list(prev_thr = ORA_res_tbl_GO$ID,
                     new_thr = ORA_res_GO_new_thr$ID)
plot(Venn(geneset_list))
```

</details>

#### Question {-}

Repeat the ORA analysis using this time the Hallmarks genesets.

<details>
<summary>Solution</summary>


```{r ora-hallmark-ex}
# Define our DE genes
padj_thr <- 1e-10
log2FC_thr <- 0

genes_DE <- res_tbl |>
    filter(padj <= padj_thr,
           abs(log2FoldChange) >= log2FC_thr) |>
    pull(gene)

geneList <- res_tbl$gene

hallmarks <- msigdbr(species = "Homo sapiens",
                     collection = "H") |>
    dplyr::select(gs_name, gene_symbol)

ORA_hallmarks <- enricher(gene          = genes_DE,
                          universe      = geneList,
                          pAdjustMethod = "BH",
                          qvalueCutoff = 1,
                          pvalueCutoff = 1,
                          TERM2GENE     = hallmarks) |>
    as_tibble()

ORA_hallmarks_tbl <- ORA_hallmarks |>
    filter(p.adjust < 0.05)

ORA_hallmarks_tbl
```

</details>

## Gene set enrichment analysis

### Principle

Gene set enrichment analysis (GSEA) refers to a broad family of
tests. Here, we will define the principles based on
[@Subramanian:2005], keeping in mind that the exact implementation
will differ in different tools.

Gene Set Enrichment Analysis (GSEA) is a statistical method used to
determine whether predefined sets of genes (for example, genes
belonging to a GO term) show systematic differences in expression
between two biological conditions.  Instead of focusing only on
individual genes, GSEA evaluates the **collective behavior of gene
groups**.

The major advantage of GSEA approaches is that they don't rely on
defining DE genes.

#### Step 1: Gene ranking {-}

The first step is to order the genes of interest. Genes could be
ranked for example by the value of the test statistic or by their
p-value.

Depending on the metric used for the ranking, genes will be ordered
from the most up-regulated to the most down-regulated (if the test
statistic is used for instance) or by the most significantly DE (no
matter if they are up or down-regulated) to genes not differentially
expressed.

#### Step 2: Check where the genes of a specific geneset appear {-}

For a given gene set (a GO-term, or a KEGG pathaway for example), the
idea is to check if these genes tend to appear toward the top or
bottom of the ranked list, rather than being randomly scattered.

#### Step 3: Compute an Enrichment Score (ES) {-}

The algorithm walks down the ranked list, increasing a running-sum
statistic when it encounters a gene from the set, and decreasing it
otherwise.

The positive score is defined by $\frac{n_{genes} -
n_{genes~in~set}}{n_{genes~in~set}}$, and the decreasing score by -1,
so that the sum of all genes in the set and those not in the set
becomes zero.

The maximum deviation from zero is the Enrichment Score (ES),
reflecting how strongly the gene set is enriched at either end of the
ranking.


```{r gsea-step-3, echo=FALSE, warning=FALSE, fig.height = 5, fig.width = 8}
geneList <- res_tbl$stat
names(geneList) <- res_tbl$gene
geneList <- geneList[order(geneList, decreasing = TRUE)]

set.seed(1)
GSEA_res <- GSEA(geneList,
             seed = TRUE,
             TERM2GENE = GO_BP,
             verbose = FALSE,
             by = "fgsea",
             pvalueCutoff = 1,
             pAdjustMethod = "BH",
             eps = 0)

GSEA_res_tbl <- as_tibble(GSEA_res) |>
    filter(p.adjust < 0.05)

gseaplot(GSEA_res,
         GSEA_res_tbl$ID[1],
         by = "runningScore",
         title = GSEA_res_tbl$ID[1])
```


#### Step 4: Assess significance by permutation {-}

To evaluate whether the observed Enrichment Score (ES) is greater than
expected by chance, GSEA performs permutations (of either gene labels
or sample labels). This generates a null distribution to compute a
p-value.

An Enrichment Score (ES) is recomputed for each permutation. The
p-value is given by the proportion of permutations where the permuted
ES is at least as extreme as the observed ES:

$$p = \frac{\#\{ ES_{\text{permuted}} \ge ES_{\text{observed}} \}}{\text{total number of permutations}}$$

### Running a GSEA with the `clusterProfiler` package

Here, we will re-use the [clusterProfiler
package](https://yulab-smu.top/clusterProfiler-book/) to run a GSEA
this time, using `hallmark genesets`.

```{r run-gsea-hallmark, message = FALSE, warning = FALSE}
# Ranking our DE results by the statistic value used in DESeq2
geneList <- res_tbl$stat
names(geneList)<- res_tbl$gene
geneList <- geneList[order(geneList, decreasing = TRUE)]

set.seed(1)
GSEA_res <- GSEA(geneList,
             TERM2GENE = hallmarks,
             verbose = FALSE,
             by = "fgsea",
             pvalueCutoff = 1,
             pAdjustMethod = "BH")

GSEA_hallmarks_tbl <- GSEA_res |>
    as_tibble() |>
    filter(p.adjust < 0.05)

GSEA_hallmarks_tbl
```

In the output data frame, there are the following columns:

- `ID`: ID of the gene set. In this example analysis, it is the GO ID.
- `Description`: Readable description. Here it is the name of the GO
  term.
- `setSize`: Number of genes in this gene set that are present in your
  universe of tested genes.
- `enrichmentScore`: Raw enrichment score calculated from the
  running-sum statistic.
- `NES`: ES normalized by gene set size.
- `pvalue`: _p_-value calculated from the hypergeometric distribution.
- `p.adjust`: Adjusted _p_-value by the BH method.
- `qvalue`: _q_-value which is another way for controlling false
  positives in multiple testings.
- `rank`: Position in the ranked gene list where the running score
  reaches its maximum (the ES).
- `leadingEdge`: List of genes in the set contributing most to the ES
  (genes before the peak), often called the core enrichment.

The GSEA plot can be drawn with the `gseaplot()` function

```{r gseaplot-hallmarks, fig.height = 5, fig.width = 8}
gseaplot(GSEA_res,
         GSEA_hallmarks_tbl$ID[1],
         by = "runningScore",
         title = GSEA_hallmarks_tbl$ID[1])
```

#### Question {-}

Compare the results obtained using ORA and GSEA on GO terms (BP)

<details>
<summary>Solution</summary>

```{r venn-ora-gsea, warning = FALSE}
library(Vennerable)
geneset_list <- list(ORA = ORA_hallmarks_tbl$ID,
                     GSEA = GSEA_hallmarks_tbl$ID)
plot(Venn(geneset_list))
```

</details>

#### Question {-}

Compare GSEA and ORA methods, what are pros and cons of both methods?

#### Question {-}

Think about the different ways to order the genes for the GSEA analysis.
What impact could the ranking method have on the GSEA analysis?
