---
title: "RNA-Seq analysis with R and Bioconductor"
author: "Laurent Gatto and Axelle Loriot"
date: "`r Sys.Date()`"
description: "This course illustrates how to understand and analyse bulk RNA-seq data, from data generation to interpretation."
output:
  bookdown::gitbook:
    config:
      download: "pdf"
      sharing:
        facebook: false
        twitter: false
        all: []
documentclass: book
link-citations: yes
bibliography: [refs.bib, packages.bib]
---

# Preamble {-}

The main objective of this course is to demonstrate how to proceed
with a complete analysis of RNA-seq data, including principal
component analysis, identification of differentially expressed genes
and gene set enrichment and over-representation analysis.

The background for these analyses includes introductions to hypothesis
testing, linear modelling and adjustment for multiple comparisons,
principal component analysis and high throughput sequencing.

This course is based on material from the
[WSBIM1322](https://uclouvain-cbio.github.io/WSBIM1322/) and
[WSBIM2122](https://uclouvain-cbio.github.io/WSBIM2122) chapters
taught at the [Faculty of Pharmacy and Biomedical
Sciences](https://uclouvain.be/en/faculties/fasb) at the UCLouvain.

## Installation instructions {-}

```{r install, eval=FALSE}
pks <- c("tidyverse", "patchwork", "remotes",
         "SummarizedExperiment", "DESeq2",
         "AnnotationDbi", "factoextra", "pRolocdata",
         "org.Hs.eg.db", "GO.db", "msigdbr", "apeglm",
         "clusterProfiler", "airway")

if (!require("BiocManager"))
    install.packages("BiocManager")

BiocManager::install(pks)
```

<!-- ## Data preparation  {-} -->

```{r prepare-wsbim2122-data, echo = FALSE, message = FALSE}
unlink("wsbim2122_data", recursive = TRUE)
rWSBIM2122::prepare_shell()
```

## References {-}

- References are provided throughout the workshop material.

- **Modern Statistics for Modern Biology**, by Susan Holmes and
  Wolfgang Huber [@MSMB]. A free online version of the book is
  available [here](https://www-huber.embl.de/msmb/).
