---
title: "RNA-Seq analysis with R and Bioconductor"
author: "Laurent Gatto and Axelle Loriot"
date: "`r Sys.Date()`"
description: "This course illustrates how to understand and analyse bulk RNA-seq data, from data generation to interpretation."
output:
  bookdown::gitbook:
    config:
      download: "pdf"
      sharing:
        facebook: false
        twitter: false
        all: []
documentclass: book
link-citations: yes
urlcolor: blue
bibliography: [refs.bib, packages.bib]
---

# Preamble {-}

RNA sequencing (RNA-Seq) is a powerful technology that is now commonly
used to measure gene expression. A differential expression analysis is
a statistical analysis that aims to identify quantitative changes in
gene expression levels between different experimental groups. The
course is aimed at biologists who are already familiar with the R
language and who want to learn how to do a differential expression
analysis on their RNA-sseq data.

The course will provide:

- Overview on RNA-seq data processing, exploring the different steps
  from raw reads to the generation of a gene count matrix.
- Introduction to theoretical concepts behind a differential
  expression analysis: linear models, multiple testing correction,
  principal component analysis, negative binomial distribution, and
  gene set enrichment analysis.
- Numerous hands-on exercices.
- A complete real-life RNA-Seq DE analysis running the [DESeq2
  package](https://bioconductor.org/packages/release/bioc/html/DESeq2.html)
  step by step.

##  Prerequisits {-}

Participants should be familiar with the basix syntax of R and ideally
also the tidyverse. They must bring a laptop with a Mac, Linux, or
Windows operating system (not a tablet, Chromebook, etc.) that they
have administrative privileges on. They should have installed R
(https://cloud.r-project.org/) and Rstudio
(https://rstudio.com/products/rstudio/download/#download).

## Authors {-}

This material has been developed by Dr Axelle Loriot and Prof Laurent
Gatto, both members of the [Computional Biology and Bioinformatics
lab](https://lgatto.codeberg.page/cbio-lab/) and the [bioinformatics
core facility](https://sites.uclouvain.be/training/bioinfo/).

This course is based on material from the
[WSBIM1322](https://uclouvain-cbio.github.io/WSBIM1322/) and
[WSBIM2122](https://uclouvain-cbio.github.io/WSBIM2122) courses taught
at the [Faculty of Pharmacy and Biomedical
Sciences](https://uclouvain.be/en/faculties/fasb) at the UCLouvain,
and the [RNA-seq analysis with
Bioconductor](https://carpentries-incubator.github.io/bioc-rnaseq/)
Carpentries workshop developed by the [Bioconductor training
committee](https://training.bioconductor.org/).

## License {-}

This book is distributed under a Attribution-NonCommercial-ShareAlike
([CC BY-NC-SA](https://creativecommons.org/licenses/by-nc-sa/4.0/))
4.0 license. This means that

You are free to:

- **Share** - copy and redistribute the material in any medium or format
- **Adapt** - remix, transform, and build upon the material
- The licensor cannot revoke these freedoms as long as you follow the
  license terms.

Under the following terms:

- **Attribution** - You must give appropriate credit, provide a link
  to the license, and indicate if changes were made. You may do so in
  any reasonable manner, but not in any way that suggests the licensor
  endorses you or your use.
- **NonCommercial** - You may not use the material for commercial
  purposes.
- **ShareAlike** - If you remix, transform, or build upon the
  material, you must distribute your contributions under the same
  license as the original.
- **No additional restrictions** - You may not apply legal terms or
  technological measures that legally restrict others from doing
  anything the license permits.

Notices:

You do not have to comply with the license for elements of the
material in the public domain or where your use is permitted by an
applicable exception or limitation.

No warranties are given. The license may not give you all of the
permissions necessary for your intended use. For example, other rights
such as publicity, privacy, or moral rights may limit how you use the
material.

## Installation instructions {-}

[Bioconductor](https://bioconductor.org/)[^about-bioc] packages can be
[installed](https://bioconductor.org/install/) with the
`BiocManager::install()` function from the `BiocManager` package, that
can be installed from CRAN with the usual `install.packages()`
function.

[^about-bioc] We will explain more about the Bioconductor projet
later, when learning about the `SummarizedExperiment` class.

```{r installation-instructions, eval=FALSE}
pks <- c("tidyverse", "patchwork", "remotes",
         "SummarizedExperiment", "DESeq2",
         "AnnotationDbi", "factoextra", "pRolocdata",
         "org.Hs.eg.db", "GO.db", "msigdbr", "apeglm",
         "clusterProfiler", "airway")

if (!require("BiocManager"))
    install.packages("BiocManager")

BiocManager::install(pks)
```

<!-- ## Data preparation  {-} -->

```{r prepare-wsbim2122-data, echo = FALSE, message = FALSE}
unlink("wsbim2122_data", recursive = TRUE)
rWSBIM2122::prepare_shell()
```

## References {-}

- **Modern Statistics for Modern Biology**, by Susan Holmes and
  Wolfgang Huber [@MSMB]. A free online version of the book is
  available [here](https://www-huber.embl.de/msmb/).
- Additional references are provided throughout the workshop material.
