---
title: "RNA-Seq analysis with R and Bioconductor"
author: "Laurent Gatto and Axelle Loriot"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
knit: bookdown::preview_chapter
# output:
#   msmbstyle::msmb_html_book:
#     toc: TRUE
#     toc_depth: 1
#     split_by: chapter
#     split_bib: no
#   css: style.css
link-citations: yes
bibliography: [refs.bib, packages.bib]
---

# Preamble {-}

The main objective of this course is to demonstrate how to proceed
with a complete analysis of RNA-seq data, including principal
component analysis, identification of differentially expressed genes
and gene set enrichment and over-representation analysis.

The background for these analyses inludes introductions to hypthesis
testing, linear modelling and adjustment for multiple comparisons,
principal component analysis and high throughtput sequencing.

This course is based on material from the
[WSBIM1322](https://uclouvain-cbio.github.io/WSBIM1322/) and
[WSBIM2122](https://uclouvain-cbio.github.io/WSBIM2122) chapters
taught at the [Faculty of Pharmacy and Biomedical
Sciences](https://uclouvain.be/en/faculties/fasb) at the UCLouvain.

## Installation instructions {-}

```{r install, eval=FALSE}
pks <- c("tidyverse", "patchwork", "remotes",
         "SummarizedExperiment", "DESeq2",
         "AnnotationDbi", "factoextra", "pRolocdata",
         "org.Hs.eg.db", "GO.db", "msigdbr", "apeglm",
         "clusterProfiler", "airway")

if (!require("BiocManager"))
    install.packages("BiocManager")

BiocManager::install(pks)
```

<!-- ## Data preparation  {-} -->

```{r prepare-wsbim2122-data, echo = FALSE, message = FALSE}
unlink("wsbim2122_data", recursive = TRUE)
rWSBIM2122::prepare_shell()
from <- list.files("wsbim2122_data/count_data", full.names = TRUE)
to <- sub("wsbim2122_data/count_", "", from)
stopifnot(all(file.rename(from, to)))
```

## References {-}

- References are provided throughout the workshop material.

- **Modern Statistics for Modern Biology**, by Susan Holmes and
  Wolfgang Huber [@MSMB]. A free online version of the book is
  available [here](https://www-huber.embl.de/msmb/).
